Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-02-19 manuscripts/reply/reply-main.tex
Adjusting formatting of reply ms.
Changing all `section*` calls to `section`. Switching to `\maketitle`
for formatting of the title page.

\section{The importance of power analysis to guide applications of \msb}
\citet{Hickerson2013} presented a power analysis of \msb under a narrow uniform
divergence-time prior of 0--1 coalescent units ago.
They found that under these prior conditions \msb can, assuming a
per-site rate of $1.92\e{-8}$ mutations per generation, detect multiple
divergence events among 18 taxa when the true divergences were random over
hundreds of thousands of generations or more.
\citet{Oaks2012} performed similar power analyses under three uniform
divergence-time priors as narrow as 0--5 coalescent units, and found the
method was able to detect multiple events among 22 when divergences were random
over millions of generations.
As recommended by \citet{Oaks2012}, it is important that investigators perform
power analyses to determine the method's power for their dataset, and decide if
\msb has sufficient temporal resolution to address their hypotheses; in the
case of the Philippines dataset, it did not.
When doing so, it is important to consider what prior conditions are
relevant to their empirical system.
A divergence-time prior of 0--5 coalescent units is quite narrow considering
that this expresses the prior belief that all 22 taxon pairs diverged within
this window.
Certainly, it is  rare for there to be enough \emph{a priori} information to
be certain that all taxa diverged within the last 4$\globalpopsize$ generations
(i.e., 0--1 coalescent units).
Also, it seems unlikely that when such prior information is available that
being able to detect multiple divergences on the scale of hundreds of thousands
of generations will add much insight about the evolutionary history of the
taxa; assuming a rate of mutation consistent with nuclear loci ($1\e{-9}$),
this translates to a temporal resolution of 3 million generations or more.

@@

\section{General thoughts on the model of \msb}
Our findings are not surprising in light of the difficult inference problem
with which \msb is faced.
To get a sense of the difficulty of this problem, we tally up all the free
parameters of the \msb model as applied to the dataset of \citet{Oaks2012}.
Under the simplest model in \msb (i.e., assuming no migration and no
intra-locus recombination), the number of parameters for each taxon pair
include:
The population sizes of the ancestral and descendant populations
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{}),
the magnitude of population contraction in each of the descendant
populations (\bottleScalar{1}{} and \bottleScalar{2}{}), the
timing of these contractions (\bottleTime{}), and the $N-1$ node heights
(coalescent times) of the gene tree that gave rise to the $N$ gene
copies sampled from both populations.
Lastly, there are between one and 22 divergence time parameters \divt{} in
the vector \divtvector.
Overall, when applying the simplest model in \msb to the dataset of
\citet{Oaks2012} with 22 taxon pairs, there are 581--602 free parameters
(depending on the number of divergence-time parameters in \divtvector).
Furthermore, under this rich model, the method is estimating the probability of
1002 divergence models \citep[i.e., the number of integer partitions of
$Y=22$;][]{Oaks2012}.

>>>>

\section{The importance of power analysis to guide applications of \texttt{msBayes}}
\citet{Hickerson2013} presented a power analysis of \msb under a narrow uniform
divergence-time prior of 0--1 coalescent units ago.
They found that under these prior conditions \msb can, assuming a
per-site rate of $1.92\e{-8}$ mutations per generation, detect multiple
divergence events among 18 taxa when the true divergences were random over
hundreds of thousands of generations or more.
\citet{Oaks2012} performed similar power analyses under three uniform
divergence-time priors as narrow as 0--5 coalescent units, and found the
method was able to detect multiple events among 22 when divergences were random
over millions of generations.
As recommended by \citet{Oaks2012}, it is important that investigators perform
power analyses to determine the method's power for their dataset, and decide if
\msb has sufficient temporal resolution to address their hypotheses; in the
case of the Philippines dataset, it did not.
When doing so, it is important to consider what prior conditions are
relevant to their empirical system.
A divergence-time prior of 0--5 coalescent units is quite narrow considering
that this expresses the prior belief that all 22 taxon pairs diverged within
this window.
Certainly, it is  rare for there to be enough \emph{a priori} information to
be certain that all taxa diverged within the last 4$\globalpopsize$ generations
(i.e., 0--1 coalescent units).
Also, it seems unlikely that when such prior information is available that
being able to detect multiple divergences on the scale of hundreds of thousands
of generations will add much insight about the evolutionary history of the
taxa; assuming a rate of mutation consistent with nuclear loci ($1\e{-9}$),
this translates to a temporal resolution of 3 million generations or more.

@@

\section{General thoughts on the model of \texttt{msBayes}}
Our findings are not surprising in light of the difficult inference problem
with which \msb is faced.
To get a sense of the difficulty of this problem, we tally up all the free
parameters of the \msb model as applied to the dataset of \citet{Oaks2012}.
Under the simplest model in \msb (i.e., assuming no migration and no
intra-locus recombination), the number of parameters for each taxon pair
include:
The population sizes of the ancestral and descendant populations
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{}),
the magnitude of population contraction in each of the descendant
populations (\bottleScalar{1}{} and \bottleScalar{2}{}), the
timing of these contractions (\bottleTime{}), and the $N-1$ node heights
(coalescent times) of the gene tree that gave rise to the $N$ gene
copies sampled from both populations.
Lastly, there are between one and 22 divergence time parameters \divt{} in
the vector \divtvector.
Overall, when applying the simplest model in \msb to the dataset of
\citet{Oaks2012} with 22 taxon pairs, there are 581--602 free parameters
(depending on the number of divergence-time parameters in \divtvector).
Furthermore, under this rich model, the method is estimating the probability of
1002 divergence models \citep[i.e., the number of integer partitions of
$Y=22$;][]{Oaks2012}.

### 2014-02-19 manuscripts/reply/reply-si-main.tex
Adjusting formatting of reply ms.
Changing all `section*` calls to `section`. Switching to `\maketitle`
for formatting of the title page.

\subsection{Differing utilities of \numt{} and \vmratio{} in \msb}
The primary component of the \msb model is the vector of divergence
times for each of the taxon pairs,
$\divtvector = \{\divt{1}, \ldots, \divt{Y}\}$
\citep{Oaks2012}.
\citet{Hickerson2013} argue that the dispersion index of this vector,
\vmratio{}, is a better model-choice estimator than the number of 
divergence-time parameters within the vector,
\numt{}.
They present a plot of \numt{} against \vmratio{} (Fig.~S1 of
\citet{Hickerson2013}), which is essentially a plot of sample size versus
variance.
This plot shows, not surprisingly, that \vmratio{} has very little information
about the number of divergences among taxa.
Nonetheless, \citet{Hickerson2013} conclude \vmratio{} is more informative and
biogeographically relevant than \numt{}.
% We struggle to follow this logic.
However, all of the information about the temporal distribution of divergences
is contained within the divergence-time vector that \vmratio{} is summarizing.
Clearly, the number of divergence time parameters within the vector and their
values is more informative than its variance (i.e., the dispersion index is not
a sufficient statistic for \divtvector).
\citet{Hickerson2013} also argue that ``\msb can estimate \vmratio{} much
better than \numt{}.''
However, \citet{Oaks2012} demonstrate that even when all assumptions of the
model are met, \vmratio{} is a poor model-choice estimator (see plots B, D \& F
of Figure 4 in \citet{Oaks2012}), whereas \numt{} performs better.

>>>>

\subsection{Differing utilities of \numt{} and \vmratio{} in \texttt{msBayes}}
The primary component of the \msb model is the vector of divergence
times for each of the taxon pairs,
$\divtvector = \{\divt{1}, \ldots, \divt{Y}\}$
\citep{Oaks2012}.
\citet{Hickerson2013} argue that the dispersion index of this vector,
\vmratio{}, is a better model-choice estimator than the number of 
divergence-time parameters within the vector,
\numt{}.
They present a plot of \numt{} against \vmratio{} (Fig.~S1 of
\citet{Hickerson2013}), which is essentially a plot of sample size versus
variance.
This plot shows, not surprisingly, that \vmratio{} has very little information
about the number of divergences among taxa.
Nonetheless, \citet{Hickerson2013} conclude \vmratio{} is more informative and
biogeographically relevant than \numt{}.
% We struggle to follow this logic.
However, all of the information about the temporal distribution of divergences
is contained within the divergence-time vector that \vmratio{} is summarizing.
Clearly, the number of divergence time parameters within the vector and their
values is more informative than its variance (i.e., the dispersion index is not
a sufficient statistic for \divtvector).
\citet{Hickerson2013} also argue that ``\msb can estimate \vmratio{} much
better than \numt{}.''
However, \citet{Oaks2012} demonstrate that even when all assumptions of the
model are met, \vmratio{} is a poor model-choice estimator (see plots B, D \& F
of Figure 4 in \citet{Oaks2012}), whereas \numt{} performs better.

### 2014-02-19 manuscripts/reply/reply.tex
Updating reply ms formating.
Changing all `section*` calls to `section`.
Adjusting formatting of reply ms.
Changing all `section*` calls to `section`. Switching to `\maketitle`
for formatting of the title page.

\newcommand{\msTitle}{Implications of uniformly distributed, empirically
    informed priors for phylogeographical model selection: A reply to Hickerson
    et al.
\xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\doublespacing
\raggedright
\setlength{\parindent}{0.5in}
\begin{linenumbers}

\begin{titlepage}
    \begin{flushleft}
        \sffamily

        \MakeUppercase{\large\bfseries \msTitle}

        \vspace{12pt}
        \textbf{Running head:} \MakeUppercase{Approximate Bayesian model
        choice}

        \vspace{12pt}
        Jamie R.\ Oaks$^{1,2}$, Jeet Sukumaran$^{3}$, Charles W.\ Linkem$^{4}$,
        Cameron D.\ Siler$^{5}$, and Rafe M.\ Brown$^{1}$

        \bigskip
        $^1$\emph{Department of Ecology and Evolutionary Biology,
            Biodiversity Institute,
            University of Kansas,
            Lawrence, Kansas 66045}\\[.1in]
        $^3$\emph{Department of Biology,
            Duke University,
            Durham, North Carolina 27708} \\[.1in]
        $^4$\emph{Department of Biology,
            University of Washington,
            Seattle, Washington 98195}\\[.1in]
        $^5$\emph{Sam Noble Museum,
            Department of Biology,
            University of Oklahoma,
            Norman, Oklahoma 73072}\\[.1in]
        $^2$\emph{Corresponding author} (\href{mailto:joaks1@gmail.com}{\tt
        joaks1@gmail.com})\\

    \end{flushleft}
\end{titlepage}

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citet{Huang2011} introduced an approximate Bayesian approach which can be
    used to estimate the probabilities of models in which multiple sets of taxa
    diverge at the same time. 
    This method (implemented in the software \msb) does not require a common
    set of character data that links all of the taxa under investigation, but
    does require the relative mutation rates among the taxa and loci to be
    assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.

    \vspace{12pt}
    \noindent\textbf{KEY WORDS: Approximate Bayesian computation; Bayesian
        model choice; empirical Bayes} 
}

@@

\section*{Acknowledgments}
We thank Jake Esselstyn for insightful comments that greatly improved this
work.
J.\ Oaks thanks the National Science Foundation for supporting this work (DEB
1011423 and DBI 1308885).
J.\ Oaks was also supported by the University of Kansas (KU) Office of Graduate
Studies, Society of Systematic Biologists, Sigma Xi Scientific Research
Society, KU Department of Ecology and Evolutionary Biology, and the KU
Biodiversity Institute.

@@

% PUT MAIN TEXT CITATION HERE
{
\renewcommand{\refname}{\noindent\MakeUppercase{\LARGE\sffamily\upshape supporting information}}
\begin{thebibliography}{1}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\urlprefix}{URL }

\bibitem[{Oaks et~al.(2014)Oaks, Sukumaran, Esselstyn, Linkem, Siler, Brown,
  and Holder}]{Oaks2013reply}
Oaks, J.~R., J.~Sukumaran, J.~A.\ Esselstyn, C.~W.\ Linkem, C.~D.\ Siler,
    R.~M.\ Brown, and M.~T.\ Holder,
  % 2013.
\newblock \msTitle
% \newblock Evolution 67:991--1010.

\end{thebibliography}
}

\doublespacing
\input{reply-si-main.tex}

>>>>

\title{Implications of uniformly distributed, empirically informed priors for
phylogeographical model selection: A reply to Hickerson et al.}

\author[1]{Jamie R.\ Oaks\thanks{Corresponding author: \href{mailto:joaks1@gmail.com}{\tt joaks1@gmail.com}}\thanks{Current address: Department of Biology, University of Washington, Seattle, Washington 98195}}
\author[2]{Jeet Sukumaran}
\author[3]{Charles W.\ Linkem}
\author[4]{Cameron D.\ Siler}
\author[1]{Rafe M.\ Brown}
\affil[1]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence, Kansas 66045}
\affil[2]{Department of Biology, Duke University, Durham, North Carolina 27708}
\affil[3]{Department of Biology, University of Washington, Seattle, Washington 98195}
\affil[4]{Sam Noble Museum, Department of Biology, University of Oklahoma, Norman, Oklahoma 73072}

% \date{\today}
\date{\parbox{\linewidth}{\centering%
    \today\endgraf\bigskip
    \textbf{Running head}: Approximate Bayesian model choice}}

\makeatletter
\let\msTitle\@title
\let\msAuthor\@author
\let\msDate\@date
\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
% \doublespacing
% \raggedright
% \setlength{\parindent}{0.5in}
% \begin{linenumbers}

\maketitle

\begin{abstract}
    Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citet{Huang2011} introduced an approximate Bayesian approach which can be
    used to estimate the probabilities of models in which multiple sets of taxa
    diverge at the same time. 
    This method (implemented in the software \msb) does not require a common
    set of character data that links all of the taxa under investigation, but
    does require the relative mutation rates among the taxa and loci to be
    assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.

    \vspace{12pt}
    \noindent\textbf{KEY WORDS: Approximate Bayesian computation; Bayesian
        model choice; empirical Bayes} 
\end{abstract}

@@

\section{Acknowledgments}
We thank Jake Esselstyn for insightful comments that greatly improved this
work.
J.\ Oaks thanks the National Science Foundation for supporting this work (DEB
1011423 and DBI 1308885).
J.\ Oaks was also supported by the University of Kansas (KU) Office of Graduate
Studies, Society of Systematic Biologists, Sigma Xi Scientific Research
Society, KU Department of Ecology and Evolutionary Biology, and the KU
Biodiversity Institute.

@@

\section*{Supporting Information}
% PUT MAIN TEXT CITATION HERE
\hangindent=1cm
Oaks, J.~R., J.~Sukumaran, C.~W.\ Linkem, C.~D.\ Siler, and R.~M.\ Brown. \msTitle.

% \doublespacing
\input{reply-si-main.tex}
